The Minimal Retroreflective Microfacet Model

### 2025-12-22 article.tex
Improve text.

\section{Back-vector Modification of Microfacet Models}

@@

To achieve retro-reflection, we modify the microfacet model by changing the microfacet normal selector from the half-vector to the \emph{back-vector}~\cite{BelcourBackvector}, defined as
\begin{equation}
  \vecbr(\vecv, \vecl) = \vechr(\vecvv, \vecl) = \frac{\vecvv + \vecl}{\lVert \vecvv + \vecl \rVert}, \text{ with } \vecvv = \text{reflect}(\vecv, \vecn),
\end{equation}
to select the reflecting microfacet. %, we are to re-derive the microfacet model.
For refraction, we can construct a similar refraction back-vector $\vecbt(\vecv, \vecl) = \vecht(\vecvv, \vecl)$.

@@

\paragraph*{Final form.}
It follows from the above back-vector modifications that our retroreflective model takes the trivial form of simply replacing $\vecv$ by $\vecvv$ in the original microfacet BRDF and BTDF expressions:
\begin{equation} \label{eq:retro_microfacet_model}
  %  f_\text{MRM}(\vecv, \vecl) = f(\vecvv, \vecl) = \frac{D(\vecb) \, G_{2}(\vecvv,\vecl) \, F(\vecvv, \vecb)}{4 \vdot{\vecvv}{\vecn} \vdot{\vecl}{\vecn}}
  f_\text{MRM}(\vecv, \vecl) =
  \begin{cases}
    f_{r, \text{MRM}}(\vecv, \vecl) = f_r(\vecvv, \vecl)  & \text{if \;} (\vdot{\vecv}{\vecn})(\vdot{\vecl}{\vecn}) \geq 0, \\
    f_{t, \text{MRM}}(\vecv, \vecl) = f_t(\vecvv, \vecl)  & \text{otherwise}.
  \end{cases}
\end{equation}
\begin{figure}[!tb]
  \centering
  \includegraphics[width=0.93\linewidth]{figures/brdf_plots.pdf}
  \caption{Comparison of the retro-reflective MRM BRDF (blue) to the regular GGX microfacet BRDF (green) and rough-diffuse EON model (orange), as a function of view direction $\theta_o$ for fixed light directions $\theta_i \in {45^\circ, 80^\circ}$. The dashed lines are the roughness $0.1$ case, and the dashed lines roughness $0.666$. The grey horizontal line at $1/\pi$ is the Lambert BRDF.}
  \label{fig:brdf_plots}
\end{figure}

@@

\begin{itemize}
\item
  Evaluation merely needs to replace $\vecv$ with $\vecvv$ upfront.
\item
  Similarly, importance sampling of $\vecl$ given $\vecv$ can be realized by replacing $\vecv$ with $\vecvv$ upfront and then importance sampling the regular microfacet BSDF.
  This may include low variance sampling using the domain of visible microfacets \cite{HeitzIS}.
\item
  As the albedos of standard BSDF and retro-reflective BSDF are identical, compensating for the energy loss of microfacet single-scattering via a diffuse term \cite{KelemenBRDF, Hill2017PBR} or normalization \cite{TurquinMultipleScattering} can thus be realized using the same data tables. The exact energy compensation of \citet{Heitz2016} can also be applied unchanged.
\item An implementation may optionally wish to provide a blend between regular and retro-reflective microfacet behavior, to model materials with partial retro-reflective properties. This can be achieved simply by blending $f_{r, \text{MRM}}$ and $f_r$, or alternatively via a stochastic selection (e.g. one-sample MIS \cite{Veach1998}) between the original $\vecv$ and reflected $\vecvv$ view directions during sampling.
\end{itemize}

%\noindent
\begin{lstlisting}[
    style=snippet,
    language=GLSL,
    caption=Pseudo code to implement the retro-reflective microfacet model using an existing regular microfacet BSDF implementation,
    label=lst:implementation,
    nolol=true,
    frame=trBL
    ]
Sample_result sample_bsdf_retro(vec3 V, vec3 L, vec3 N, Bsdf_params params)
{
    vec3 Vp = -V + 2.0 * dot(V, N) * N;
    return sample_bsdf(Vp, L, N, params);
}
Eval_result eval_bsdf_retro(vec3 V, vec3 L, vec3 N, Bsdf_params params)
{
    vec3 Vp = -V + 2.0 * dot(V, N) * N;
    return eval_bsdf(Vp, L, N, params);
}
\end{lstlisting}

@@

The albedo is identical for forward- and retro-reflective BRDF as well as BTDF, which allows to arbitrarily combine forward- and retro-reflective variants into a BSDF.
When constructing an artist-controllable retro-reflective dielectric BSDF, we observed that using $f_t$ instead of $f_{t, \text{MRM}}$ yields more intuitive results, in particular for an index of refraction close to 1.

>>>>

\section{MRM as Back-vector Modification of Microfacet Models}

@@

To achieve retro-reflection, in MRM we modify the microfacet model by changing the microfacet normal selector from the half-vector to the \emph{back-vector}~\cite{BelcourBackvector}, defined as
\begin{equation}
  \vecbr(\vecv, \vecl) = \vechr(\vecvv, \vecl) = \frac{\vecvv + \vecl}{\lVert \vecvv + \vecl \rVert}, \text{ with } \vecvv = \text{reflect}(\vecv, \vecn),
\end{equation}
to select the reflecting microfacet. %, we are to re-derive the microfacet model.
For refraction, we can construct a similar refraction back-vector $\vecbt(\vecv, \vecl) = \vecht(\vecvv, \vecl)$.

@@

\paragraph*{Final form.}
It follows from the above back-vector modifications that our MRM retroreflective model takes the trivial form of simply replacing $\vecv$ by $\vecvv$ in the original microfacet BRDF and BTDF expressions:
\begin{equation} \label{eq:retro_microfacet_model}
  %  f_\text{MRM}(\vecv, \vecl) = f(\vecvv, \vecl) = \frac{D(\vecb) \, G_{2}(\vecvv,\vecl) \, F(\vecvv, \vecb)}{4 \vdot{\vecvv}{\vecn} \vdot{\vecl}{\vecn}}
  f_\text{MRM}(\vecv, \vecl) =
  \begin{cases}
    f_{r, \text{MRM}}(\vecv, \vecl) = f_r(\vecvv, \vecl)  & \text{if \;} (\vdot{\vecv}{\vecn})(\vdot{\vecl}{\vecn}) \geq 0, \\
    f_{t, \text{MRM}}(\vecv, \vecl) = f_t(\vecvv, \vecl)  & \text{otherwise}.
  \end{cases}
\end{equation}
\begin{figure}[!tb]
  \centering
  \includegraphics[width=0.93\linewidth]{figures/brdf_plots.pdf}
  \caption{Comparison of the retro-reflective MRM BRDF (blue) to the regular GGX microfacet BRDF (green) and rough-diffuse EON model (orange), as a function of view direction $\theta_o$ for fixed light directions $\theta_i \in {45^\circ, 80^\circ}$. The dashed lines are the roughness $0.1$ case, and the dashed lines roughness $0.666$. The grey horizontal line at $1/\pi$ is the Lambert BRDF.}
  \label{fig:brdf_plots}
\end{figure}

@@

\begin{itemize}
\item
  Evaluation merely needs to replace $\vecv$ with $\vecvv$ upfront.
\item
  Similarly, importance sampling of $\vecl$ given $\vecv$ can be realized by replacing $\vecv$ with $\vecvv$ upfront and then importance sampling the regular microfacet BSDF.
  This may include low variance sampling using the domain of visible microfacets \cite{HeitzIS}.
\item
  As the albedos of standard BSDF and retro-reflective BSDF are identical, compensating for the energy loss of microfacet single-scattering via a diffuse term \cite{KelemenBRDF, Hill2017PBR} or normalization \cite{TurquinMultipleScattering} can thus be realized using the same data tables. The exact energy compensation of \citet{Heitz2016} can also be applied unchanged.
\item An implementation may optionally wish to provide a blend between regular and retro-reflective microfacet behavior, to model materials with partial retro-reflective properties (for example this is required in OpenPBR \cite{OpenPBR2024}). This can be achieved simply by blending $f_{r, \text{MRM}}$ and $f_r$, or alternatively via a stochastic selection (e.g. one-sample MIS \cite{Veach1998}) between the original $\vecv$ and reflected $\vecvv$ view directions during sampling.
\end{itemize}

%\noindent
\begin{lstlisting}[
    style=snippet,
    language=GLSL,
    caption=Pseudo code to implement the retro-reflective microfacet model using an existing regular microfacet BSDF implementation,
    label=lst:implementation,
    nolol=true,
    frame=trBL
    ]
Sample_result sample_bsdf_retro(vec3 V, vec3 L, vec3 N, Bsdf_params params)
{
    vec3 Vp = -V + 2.0 * dot(V, N) * N;
    return sample_bsdf(Vp, L, N, params);
}

Eval_result eval_bsdf_retro(vec3 V, vec3 L, vec3 N, Bsdf_params params)
{
    vec3 Vp = -V + 2.0 * dot(V, N) * N;
    return eval_bsdf(Vp, L, N, params);
}
\end{lstlisting}

@@

The albedo is identical for forward- and retro-reflective BRDF as well as BTDF, which allows to arbitrarily combine forward- and retro-reflective variants into a BSDF. When constructing an artist-controllable retro-reflective dielectric BSDF, we observed that using $f_t$ instead of $f_{t, \text{MRM}}$ yields more intuitive results, in particular for an index of refraction close to 1. An implementation could reasonably choose either this formulation or the full retro-reflective BTDF, depending on the desired effect.
